PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-04-21 papers/NAR_web_server_2009/paper.sup.tex
Final version of supplementary materials

Once the MST is has been calculated, our last step is to remove all
edges exceeding a threshold weight $T$, taking the resulting
components as our clusters.  \textsc{wcd}, when faced with a similar
challenge, sets the threshold at $T=40$, hypothesizing that EST pairs
with a $d^2$ distance of greater than 40 are unlikely to overlap.  \mc{We
have set our threshold to the same $T=40$ (with discussion below on
the effects of changing this).}

@@

\mc{By restricting $d^2$ comparisons to within a group, we minimize the
effects of the heterogeneous sequences lengths by restricting our
comparisons to only sequences of similar lengths.  By not restricting
ourselves to a single window size, we allow ourselves to, when possible, avoid the
performance hit of using a small window.  As we are not comparing
sequences from different groups we bypass the question of how to
compare a long sequence against a short one.  And because the
groups overlap in content, we still get implicit comparisons between
long and short sequences in terms of the resulting path of the minimum
spanning tree that makes use of these ``linking'' nodes.}

\mc{In the \textsc{peace} implementation, we subdivide into three groups:
\begin{itemize}
\item Less than 150 ($w=50$, $T=50$, $t=20$, $u=4$).
\item Between 150 and 400 ($w=75$, $T=75$, $t=35$, $u=6$)
\item Greater than 400 ($w=100$, $T=100$, $t=50$, $t=50$).
\end{itemize}}

@@

\mc{Our primary measurement tools if measurement are sensitivity, the
Jaccard Index, Type 1 error and Type 2 error
(Figure~3).  Sensitivity is the ratio for true
positives to actual positives ($tp / (tp + fn)$) -- the fraction of
sequences from the same transcript that are identified as such.  The
{\it Jaccard} index combines both sensitivity and specificity,
calculated as the ratio of true positives to the sum of true
positives, true negatives, and false positives ($tp / (tp+fn+fp)$)
\cite{Hazelhurst08a}.  Notably, when we removed duplicated genes
before generating the simulated EST sets, we found all three tools to
have perfect specificity: no false positives.}

@@

\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.35]{pics.d/duplicates_40.pdf}
}
\caption{Ability to distinguish duplicates as a function of
  divergence.  Estimates averaged over 30 trials; one trial consists
  of taking a random gene zebra-fish gene, copying it and stochastically changing
  bases at the specified rate, then using the two genes as the bases
  for generating a simulated set. Blue/Solid = \peace, Green/Dashed =
  \wcd\/ (mostly obscured by the blue line), Black/Dot-Dashed =
  \capthree; variance was too small for visible plotting of
  confidence intervals.}\label{dups}
\end{figure}

\subsection{Raising the Threshold}
\label{sim_thresh}
\mc{ 
  The \wcd\/ tool uses a $d^2$ threshold of $40$ to determine whether
  two sequences are in the same cluster.  As this is equivalent to our
  $T$ threshold for dismissing MST edges, we have set it to the same
  default value when used for Sanger sequences.  However, we did
  experiment with higher values, to interesting results.  A basic
  analysis indicated that a $T$ value as high as 130 should not have a
  significant impact on its effectiveness as a discriminator.  And in
  fact this was born out in simulation, as shown in
  Figure~\label{SiJiT1T2130}.  In contrast with the equivalent results
  for a threshold of $T=40$, as shown in the manuscript, we see
  considerable improvement in \peace\/ over \wcd\/ in all metrics
  except Type 2 error, in which \wcd\/ slightly
  better. Unsurprisingly, this also makes a significant difference in
  the tool's ability to distinguish duplications: we see fairly poor
  performance in Figure~\ref{dups130}.  
}

\mc{
  Most importantly, while the higher threshold appeared to be better
  in theory and in analysis, it did not hold in practice.  In
  Table~\ref{SangerReal130} we show the match to the manuscript
  Table~1, and observe the poor performance of
  \peace\/ when looking at the Jaccard index.  It is clear that in the
  real data, the level of similarity is considerably higher then in the
  simulations, thus making the reduced specificity of the higher
  threshold considerably more of a problem.
}

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x130.pdf}}
  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
    Type 1 error using the elevated threhold ($T=130$).  Based on the average
    over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
    fish genes (see Supplementary Materials, Section C, for more
    details).  Blue/Solid = \peace, Green/Dash = \wcd, Black/Dot-Dash
    = \capthree; vertical tics = 95\% confidence intervals on
    estimates.  Intervals are not presented for Type 1 error due to
    the effective lack of variance.}\label{SeJiT1T2130}
\end{figure}

\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.35]{pics.d/duplicates_130.pdf}
}
\caption{Ability to distinguish duplicates as a function of
  divergence.  Estimates averaged over 30 trials; one trial consists
  of taking a random zebra-fish gene, copying it and stochastically changing bases
  at the specified rate, then using the two genes as the bases for
  generating a simulated set. Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3};
  variance was too small for visible plotting of
  confidence intervals.}\label{dups130}
\end{figure}

@@

\begin{table}[tbp]
\begin{center}
\begin{tabular} {|c|| c | c || c | c |}
\hline
 Benchmark & \begin{tabular}{c} Number of \\ Sequences \end{tabular}
 & \begin{tabular}{c} Number of \\ Bases \end{tabular} & Peace Runtme (s)&
 WCD Runtime (s) \\
\hline
\hline
 Human Cluster & 17733 & 11.7 Mb & 293 & 804 \\
\hline
 Cotton & 29,995 & 17.1 Mb & 222 & 339 \\
\hline
 Ricinus & 57,690 & 40.9 Mb & 1344 & 2356 \\
\hline
 A076941 & 76,941 & 32.8 Mb & 1166 & 966 \\
\hline
\hline
 A686903 & 686,904 & 294.7 Mb & 9449 & 2481 \\
\hline
\end{tabular}
\end{center}
\caption{Runtime for \peace\/ and \wcd\/ on benchmarks from the
  WCD and EasyCluster studies \cite{Hazelhurst08a, Picardi09}.
  Runtimes for the first four sets are sequential, while that last was
  run over thirty processors.}\label{RunTime}
\end{table}

@@

\mc{ In Table~\ref{SangerReal130} we see the effect on the application
  of \peace\/ to the data sets from Table~1 when
  raising the threshold $T$ to 130 (see previous discussion in
  Section~\ref{sim_thresh}).  While doing this appeared to be fairly
  effective in simulation, it did not hold up in practice.  We see
  essentially no change in the sensitivity, but significant changes in
  the Jaccard Index (and a corresponding decrease in the total number
  of clusters).  We suspect that this is due to a higher number of
  more closely related genes than are present in the simulated
  data. Regardless, without considerably more investigation, these
  results indicated we should be leaving $T$ at a default rate of 40.
}

@@

\begin{table}[tbp]
\begin{center}
\begin{tiny}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline   & \peace & 0.871 & 0.809 & 0.186 & 0.049 & 129 & 4 & 129 \\
\cline{2-9}  MetaSim 454 & \wcd(1) & 0.073 & 0.073 & 0.014 & 0.082 & 7028.500 & 5296.533 & 74 \\
\cline{2-9}  Simulation & \wcd(2) &  0.020 & 0.020 & 0.008 & 0.198 & 13085 & 10610 & 64 \\
\cline{2-9}  & \capthree & 0.005 & 0.005 & 0.005 & 0.565 & 19222 & 15425 & 132 \\
\hline
\hline Metasim  & \peace & 0.583 & 0.372 & 0.230 & 0.111 & 392.914 & 149.257 & 3454 \\
\cline{2-9} Illumina  & \capthree & 0.006 & 0.006 & 0.002 & 1.766 & 51663.057 & 30107.429 & 291 \\
\hline
\end{tabular}
\end{tiny}
\end{center}
\caption{Comparisons of runs on on short read data, as generated by
  \metasim \cite{Richter2008}; each figure is an average over 30 runs.
  For \wcd(1) we use wcd default parameters, while for \wcd(2) we set
  it to match the \peace\/ parameters ($w=75$, $T=60$); \wcd\/ was
  unable to return any results for Illumina data, hence has been
  omitted.  Each 454 simulation produced 25,000 reads from the 100
  zebra-fish transcripts, ranging  in size from 200bp to 270bp (aveage = 250bp,
  standard deviation = 17bp).  Each Illumina simulation produce 150K
  reads from the 100 zebra-fish transcript, each exactly 62bp.}\label{SRSim}
\end{table}

>>>>

Once the MST has been calculated, our last step is to remove all
edges exceeding a threshold weight $T$, taking the resulting
components as our clusters.  \wcd, when faced with a similar
challenge, sets the threshold at $T=40$, hypothesizing that EST pairs
with a $d^2$ distance of greater than 40 are unlikely to overlap.  \mc{We
have set our threshold to the same $T=40$ (with discussion below on
the effects of changing this).}

@@

\mc{By restricting $d^2$ comparisons to within a group, we minimize
  the effects of the heterogeneous sequences lengths by restricting
  our comparisons to only sequences of similar lengths.  By not
  restricting ourselves to a single window size, we allow ourselves
  to, when possible, avoid the performance hit of using a small
  window.  As we are not comparing sequences from groups corresponding
  to radically different sizes we bypass the question of how to
  compare a long sequence against a short one.  And because we allow
  for the comparison of sequences in neighboring groups, we still get
  implicit comparisons between long and short sequences in terms of
  the resulting path of the minimum spanning tree that makes use of
  these ``linking'' nodes.}

\mc{In the \textsc{peace} implementation, we subdivide into three groups:
\begin{itemize}
\item Less than 150bp ($w=50$, $T=50$, $t=20$, $u=4$).
\item Between 150bp and 400bp ($w=75$, $T=75$, $t=35$, $u=6$)
\item Greater than 400bp ($w=100$, $T=100$, $t=50$, $t=50$).
\end{itemize}}

@@

\mc{Our primary measurement tools if measurement are sensitivity, the
Jaccard Index, Type 1 error and Type 2 error
(Figure~3).  Sensitivity is the ratio for true
positives to actual positives ($tp / (tp + fn)$) -- the} \mc{fraction} \mc{of}
\mc{sequences} \mc{from} \mc{the} \mc{same} \mc{transcript} \mc{that} \mc{are} \mc{identified} \mc{as} \mc{such.}  \mc{The}
\mc{{\it Jaccard}} \mc{index} \mc{combines} \mc{both} \mc{sensitivity} \mc{and} \mc{specificity,}
\mc{calculated} \mc{as} \mc{the} \mc{ratio} \mc{of} \mc{true} \mc{positives} \mc{to} \mc{the sum of true
positives, true negatives, and false positives ($tp / (tp+fn+fp)$)
\cite{Hazelhurst08a}.}

@@

\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.35]{pics.d/duplicates_40.pdf}
}
\caption{Ability to distinguish duplicates as a function of
  divergence.  Estimates averaged over 30 trials; one trial consists
  of taking a random gene zebra fish gene, copying it and stochastically changing
  bases at the specified rate, then using the two genes as the bases
  for generating a simulated set. Blue/Solid = \peace, Green/Dashed =
  \wcd\/ (mostly obscured by the blue line), Black/Dot-Dashed =
  \capthree; variance was too small for visible plotting of
  confidence intervals.}\label{dups}
\end{figure}

\subsection{Runtime and Memory Footprint}

\mc{Runtime of the tool is analyzed in Figure~4 of the main paper.  In
Figure~\ref{memory} we present a plot of the memory footprint of each tool
as a function of input size.  Data points were generated by the application of
each tool to simulated sets of a specified size (generated by ESTSim),
measuring the maximum amount of memory required for the run.  We see that while
all tools require memory linear in the input size, \wcd\/ requires
slightly less than does \peace, while both require significantly less
memory than \capthree.  In general we find the footprint of \peace\/
to be relativily small, as compared to other clustering tools
(discussed in {\it Hazelhurst et al.} \cite{Hazelhurst08a}).}

\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.35]{pics.d/memory.pdf}
}
\caption{Plots of memory footprint against sequence set size for a
  single processor run on simulated data for each tool: \peace\/
  (blue), \wcd\/ (green) and \capthree\/ (balck), along with the
  lesat-squares fit for each.  For each fit we have an associated
  regression coefficient of $r > 0.99$.}\label{memory}
\end{figure}

\subsection{Raising the Threshold}
\label{sim_thresh}
\mc{ 
  The \wcd\/ tool uses a $d^2$ threshold of $40$ to determine whether
  two sequences are in the same cluster.  As this is equivalent to our
  $T$ threshold for dismissing MST edges, we have set it to the same
  default value when used for Sanger} \mc{sequences.}  \mc{However,} \mc{we} \mc{did}
  \mc{experiment} \mc{with} \mc{higher} \mc{values,} \mc{with} \mc{interesting} \mc{results.  An
  analysis of the preformance of $d^2$ indicated that a $T$ value as high as 130 should not have a
  significant impact on its effectiveness as a discriminator.  And in
  fact this was born out in simulation, as shown in
  Figure~\ref{SeJiT1T2130}.  In contrast with the equivalent results
  for a threshold of $T=40$, as shown in the manuscript, we see
  considerable improvement in \peace\/ over \wcd\/ in all metrics
  except Type 2 error, in which \wcd\/ slightly
  better. Unsurprisingly, this also makes a significant difference in
  the tool's ability to distinguish duplications: we see fairly poor
  performance in Figure~\ref{dups130}.  
}

\mc{ Most importantly, while the higher threshold appeared to be
  better in theory and in analysis, the result did not hold in
  practice.  In Table~\ref{SangerReal130} we show the figures
  corresponding to the manuscript Table~1 when raising the threshold
  to $130$, and observe the poor performance of \peace\/ when looking
  at the Jaccard index.  It is clear that in the real data, the level
  of transcript similarity is considerably higher then in the simulations, thus
  making the reduced specificity of the higher threshold considerably
  more of a problem.  }

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x130.pdf}}
  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
    Type 1 error using the elevated threhold ($T=130$), providing a
    contrast against Figure 3 ($T=40$).  Based on the average over 30
    simulated Sanger Sequence ESTs sets derived from 100 zebra fish
    genes.  Blue/Solid = \peace, Green/Dash = \wcd, Black/Dot-Dash =
    \capthree; vertical tics = 95\% confidence intervals on estimates.
    Intervals are not presented for Type 1 error due to the effective
    lack of variance.}\label{SeJiT1T2130}
\end{figure}

\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.35]{pics.d/duplicates_130.pdf}
}
\caption{Ability to distinguish duplicates as a function of
  divergence.  Estimates averaged over 30 trials; one trial consists
  of taking a random zebra fish gene, copying it and stochastically changing bases
  at the specified rate, then using the two genes as the bases for
  generating a simulated set. Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3};
  variance was too small for visible plotting of
  confidence intervals.}\label{dups130}
\end{figure}

@@

\begin{table}[tbp]
\begin{center}
\begin{tabular} {|c|| c | c || c | c |}
\hline
 Benchmark & \begin{tabular}{c} Number of \\ Sequences \end{tabular}
 & \begin{tabular}{c} Number of \\ Bases \end{tabular} & Peace Runtme (s)&
 WCD Runtime (s) \\
\hline
\hline
 Human Cluster & 17,733 & 11.7 Mb & 293 & 804 \\
\hline
 Cotton & 29,995 & 17.1 Mb & 222 & 339 \\
\hline
 Ricinus & 57,690 & 40.9 Mb & 1344 & 2356 \\
\hline
 A076941 & 76,941 & 32.8 Mb & 1166 & 966 \\
\hline
\hline
 A686903 & 686,904 & 294.7 Mb & 9449 & 2481 \\
\hline
\end{tabular}
\end{center}
\caption{Runtime for \peace\/ and \wcd\/ on benchmarks from the
  WCD and EasyCluster studies \cite{Hazelhurst08a, Picardi09}.
  Runtimes for the first four sets are sequential, while that last was
  run over thirty processors.}\label{RunTime}
\end{table}

@@

\mc{ In Table~\ref{SangerReal130} we see the effect on the application
  of \peace\/ to the data sets from Table~1 when
  raising the threshold $T$ to 130 (see previous discussion in
  Section~\ref{sim_thresh}).  While raising $T$ is  
  effective in simulations, it did not hold up in practice.  We see
  essentially no change in the sensitivity, but significant changes in
  the Jaccard Index (and a corresponding decrease in the total number
  of clusters).  We suspect that this is due to a higher number of
  more closely related genes than are present in the simulated
  data. Regardless, without considerably more investigation, these
  results indicated we should be leaving $T$ at a default rate of 40.
}

@@

\mc{In Table~\ref{SRSim} we present the results of application of the
tools to \metasim\/ generated simulated short read data, as discussed in
the results section of the paper.}

\begin{table}[tbp]
\begin{center}
\begin{tiny}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline   & PEACE & 0.871 & 0.809 & 0.186 & 0.049 & 129 & 4 & 129 \\
\cline{2-9}  MetaSim 454 & WCD(1) & 0.073 & 0.073 & 0.014 & 0.082 & 7028.500 & 5296.533 & 74 \\
\cline{2-9}  Simulation & WCD(2) &  0.020 & 0.020 & 0.008 & 0.198 & 13085 & 10610 & 64 \\
\cline{2-9}  & Cap3 & 0.005 & 0.005 & 0.005 & 0.565 & 19222 & 15425 & 132 \\
\hline
\hline Metasim  & PEACE & 0.583 & 0.372 & 0.230 & 0.111 & 392.914 & 149.257 & 3454 \\
\cline{2-9} Illumina  & Cap3 & 0.006 & 0.006 & 0.002 & 1.766 & 51663.057 & 30107.429 & 291 \\
\hline
\end{tabular}
\end{tiny}
\end{center}
\caption{Comparisons of runs on on short read data, as generated by
  \metasim \cite{Richter2008}; each figure is an average over 30 runs.
  For \wcd(1) we use wcd default parameters, while for \wcd(2) we set
  it to match the \peace\/ parameters ($w=75$, $T=60$); \wcd\/ was
  unable to return any results for Illumina data, hence has been
  omitted.  Each 454 simulation produced 25,000 reads from the 100
  zebra-fish transcripts, ranging  in size from 200bp to 270bp (aveage = 250bp,
  standard deviation = 17bp).  Each Illumina simulation produce 150K
  reads from the 100 zebra-fish transcript, each exactly 62bp.}\label{SRSim}
\end{table}
